---
bibliography: 
  - "references.bib"
  - "references_other.yaml"
output: rtf_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# TO CONVERT TO PLAIN TEXT:
# 1. Copy the contents of the rtf file, paste to txt.
# 2. Replace all single line breaks with double line breaks (in TextWrangler)
# 3. Hard-wrap all text at 80 characters (in Atom with autoflow: cmd-alt-q)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE,
  warning = FALSE, results = "hide", cache = FALSE)

# Load objects build during analysis from drake cache
loadd(
  list = c(
    "ppgi",
    "nectandra_specimens",
    "nectandra_dna",
    "nectandra_rbcL_raw",
    "nectandra_rbcL",
    "cyatheaceae_seqs",
    "cyatheaceae_tree",
    "grammitid_seqs",
    "grammitid_tree",
    "cr_richness",
    "iqtree_log",
    "rbcL_tree"),
  cache = nectandra_cache)

```

This README.txt file was generated on format `r format(Sys.time(), '%d %B, %Y')` by Joel Nitta

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\- GENERAL INFORMATION \-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

Title of Dataset: Data from: A taxonomic and molecular survey of the
pteridophytes of the Nectandra Cloud Forest Reserve, Costa Rica

Author Information

Principal Investigator: Joel H. Nitta

Department of Botany, National Museum of Natural History, Smithsonian
Institution, Washington, D.C., 20013, U.S.A. joelnitta@gmail.com

Associate or Co-investigator: Atsushi Ebihara  

Department of Botany, National Museum of Nature and Science, 4-1-1 Amakubo,
Tsukuba 305-0005, Japan ebihara@kahaku.go.jp

Associate or Co-investigator: Alan R. Smith

University Herbarium, University of California, Berkeley. 1001 Valley Life
Sciences Bldg. #2465. Berkeley, California, 94720, U.S.A.
arsmith@berkeley.edu

Date of data collection: 2008--2018

Geographic location of data collection: Nectandra Cloud Forest Reserve, Costa
Rica

Information about funding sources or sponsorship that supported the
collection of the data: Funding provided in part by the Nectandra Institute
and Japan Society for the Promotion of Science (Kakenhi grant no. 15K07204)

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

SHARING/ACCESS INFORMATION

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

Licenses/restrictions placed on the data, or limitations of reuse: CC0 1.0
Universal (CC0 1.0)

Recommended citation for the data: Nitta JH, Ebihara A, Smith AR (2020) Data
from: A taxonomic and molecular survey of the pteridophytes of the Nectandra
Cloud Forest Reserve, Costa Rica. Dryad Digital Repository. FIXME: ADD DOI
WHEN AVAILABLE

Citation for and links to publications that cite or use the data: Nitta JH,
Ebihara A, Smith AR (2020) A taxonomic and molecular survey of the
pteridophytes of the Nectandra Cloud Forest Reserve, Costa Rica. PLoS ONE.
FIXME: ADD DOI WHEN AVAILABLE

Code for analyzing the data is available on github:
https://github.com/joelnitta/nectandra

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

DATA & FILE OVERVIEW

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

File list (filenames, directory structure (for zipped files) and brief
description of all data files):

  - costa_rica_richness.csv: Data on species richness of pteridophytes in protected
areas in Costa Rica.

  - cyatheaceae_rbcL.fasta: Aligned rbcL sequences of family Cyatheaceae from the Nectandra Cloud Forest Reserve, Costa Rica and all available sequences on GenBank in FASTA format.

  - cyatheaceae_rbcL.tre: Phylogenetic tree of family Cyatheaceae from the Nectandra Cloud Forest Reserve, Costa Rica and all available rbcL sequences on GenBank in Newick format.

  - grammitidoideae_rbcL.fasta: Aligned rbcL sequences of subfamily Grammitidoideae from the Nectandra Cloud Forest Reserve, Costa Rica and all available sequences on GenBank in FASTA format.

  - grammitidoideae_rbcL.tre: Phylogenetic tree of subfamily Grammitidoideae from the Nectandra Cloud Forest Reserve, Costa Rica and all available rbcL sequences on GenBank in Newick format.

  - nectandra_DNA_accessions.csv: DNA accession numbers and specimen accession numbers
of pteridophytes from the Nectandra Cloud Forest Reserve, Costa Rica.

  - nectandra_rbcL.fasta: Newly generated rbcL sequences of pteridophytes from the Nectandra Cloud Forest Reserve, Costa Rica in FASTA format.

  - nectandra_rbcL.phy: Aligned rbcL sequences of pteridophytes from the Nectandra Cloud Forest Reserve, Costa Rica in PHYLIP format.

  - nectandra_rbcL.treefile: Phylogenetic tree of pteridophytes from the Nectandra Cloud Forest Reserve, Costa Rica in Newick format.

  - nectandra_specimens.csv: Specimen data of pteridophytes from the Nectandra Cloud Forest Reserve, Costa Rica collected by Joel Nitta.

  - ppgi_taxonomy.csv: Taxonomic system of Pteridophyte Phylogeny Group I (2016) for
pteridophytes at the genus level and above.

Additional related data collected that was not included in the current data
package:

  - rbcL_clean_sporos.fasta: rbcL sequences of pteridophytes of Moorea, French Polynesia [@Nitta2017; https://doi.org/10.5061/dryad.df59g].

  - ESM1.csv: A list of native fern and lycophyte taxa (species, subspecies and varieties; 721 taxa total) in Japan [@Ebihara2019b; https://doi.org/10.5061/dryad.4362p32].

  - FernGreenListV1.01E.xls: List of Japanese ferns and lycophytes species including
scientific name, endemic status, conservation status, and other taxonomic data [@Ebihara2019b; https://doi.org/10.5061/dryad.4362p32].

  - rbcl_mrbayes.nex: NEXUS file used for phylogenetic analysis of Japanese fern
and lycophyte taxa with MrBayes [@Ebihara2019b; https://doi.org/10.5061/dryad.4362p32].

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

METHODOLOGICAL INFORMATION

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

Description of methods used for collection/generation of data:

Surveys of pteridophytes (i.e., ferns and lycophytes) were carried out over
three field seasons (January 2008, 2011, and 2013;
`r nectandra_specimens %>% filter(!is.na(date_collected)) %>% pull(date_collected) %>% n_distinct` days total)
at the Nectandra Cloud Forest Reserve, Costa Rica. Most specimens were collected
along trails through the reserve. Epiphytes were collected from fallen trees
or tree branches, or up to 2 m on tree trunks. Permits for collection were
obtained from the Costa Rican government (SINAC No. 04941 and Cites 2014-CR
1006/SJ (#S 1045)). The first set of voucher specimens was deposited at UC,
with duplicates at CR and GH. Herbarium codes follow @Thiers2020. Leaf tissue
was preserved on silica gel for DNA extraction. Spores of selected taxa were
observed with a standard compound light microscope.

DNA was extracted with the DNEasy plant mini kit following the manufacturer's
protocol (Qiagen). One species per taxon was sampled for morphologically
distinct taxa, and up to five specimens per taxon for taxa that are more
difficult to identify using standard keys and morphological characters. The
plastid rbcL gene was amplified using PCR primers and thermocycler settings
of @Schuettpelz2007. PCR products were purified with Exo-STAR enzyme (GE
Healthcare) and sequenced using the Big Dye Terminator v3.1 Cycle Sequencing
Kit (ThermoFisher) with two internal primers, ESRBCL654R and ESRBCL628F
[@Schuettpelz2007] in addition to the amplification primers. The resulting
AB1 trace files were imported into Geneious [@Kearse2012], assembled into
contigs, and the consensus sequences exported in FASTA format. A
multi-sequence alignment was generated using MAFFT [@Katoh2002], and a
phylogenetic tree inferred using IQ-TREE with automatic model selection
[@Nguyen2015]. For a small number of genera that were not supported as
monophyletic in the original phylogenetic analysis (Cyathea and Lellingeria),
all available *rbcL* sequences for closely related taxa (at the family or
subfamily level, respectively) were downloaded from GenBank, aligned in
combination with the newly generated sequences from Nectandra with MAFFT, and
a phylogenetic tree inferred using FastTree on default settings [@Price2009;
@Price2010].

Molecular analysis was authorized by Convenio Marco de CONAGEBIO
R-CM-RN-001-2014-OT.

For additional methodological details, see Nitta JH, Ebihara A, Smith AR
(2020).

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

DATA-SPECIFIC INFORMATION

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

costa_rica_richness.csv: Data on species richness of pteridophytes in
protected areas in Costa Rica. Compiled by Joel Nitta based on references in
the  "citation" column.

Number of variables: `r ncol(cr_richness)`

Number of cases/rows: `r nrow(cr_richness)`

Variable list:

  - name: Abbreviated name of site.
  
  - full_name: Full name of site.

  - min_el_m: Minimum elevation of site in meters.

  - max_el_m: Maximum elevation of site in meters.

  - area_ha: Area of site in hectares.

  - richness: Number of species occurring at the site.

  - holdridge_type: @Holdridge1967 life-zone type.

  - citation: Reference for data.

  - notes: Miscellaneous notes.

  - latitude: Latitude in decimal-degrees.

  - longitude: Longitude in decimal-degrees.

Missing data codes: Missing data have no values (nothing entered between commas
in the CSV file).

Specialized formats or other abbreviations used: none.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

cyatheaceae_rbcL.fasta: Aligned rbcL sequences of family Cyatheaceae from the
Nectandra Cloud Forest Reserve, Costa Rica and all available sequences on
GenBank in FASTA format. Species from Nectandra in family Dicksoniaceae included 
as outgroup. Sequences aligned using MAFFT [@Katoh2002]. Numbers after
species names are GenBank accession numbers for sequences downloaded from
GenBank or J. H. Nitta specimen collection numbers for sequences newly
obtained by this study. `r dim(cyatheaceae_seqs)[[1]]` sequences;
`r dim(cyatheaceae_seqs)[[2]]` bp; `r ips::pis(cyatheaceae_seqs, "absolute")`
parsimony-informative sites.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

cyatheaceae_rbcL.tre: Phylogenetic tree of family Cyatheaceae from the
Nectandra Cloud Forest Reserve, Costa Rica and all available rbcL sequences
on GenBank in Newick format. Species from Nectandra in family Dicksoniaceae 
included as outgroup. Tree inferred using FastTree [@Price2009; @Price2010].
Numbers after species names are GenBank accession numbers for sequences
downloaded from GenBank or J. H. Nitta specimen collection numbers for
sequences newly obtained by this study. Numbers at nodes indicate local
support values computed with the Shimodaira–Hasegawa test. `r ape::Ntip(cyatheaceae_tree)`
tips; `r ape::Nnode(cyatheaceae_tree)` internal nodes.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

grammitidoideae_rbcL.fasta: Aligned rbcL sequences of subfamily
Grammitidoideae from the Nectandra Cloud Forest Reserve, Costa Rica and all
available sequences on GenBank in FASTA format. Sequences aligned using MAFFT
[@Katoh2002]. Species from Nectandra in subfamily Polypodioideae included as
outgroup. Numbers after species names are GenBank accession numbers for
sequences downloaded from GenBank or J. H. Nitta specimen collection numbers
for sequences newly obtained by this study. `r dim(grammitid_seqs)[[1]]` sequences;
`r dim(grammitid_seqs)[[2]]` bp; `r ips::pis(grammitid_seqs, "absolute")`
parsimony-informative sites.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

grammitidoideae_rbcL.tre: Phylogenetic tree of subfamily Grammitidoideae from
the Nectandra Cloud Forest Reserve, Costa Rica and all available rbcL
sequences on GenBank in Newick format. Species from Nectandra in subfamily
Polypodioideae included as outgroup. Tree inferred using FastTree
[@Price2009; @Price2010]. Numbers after species names are GenBank accession numbers for
sequences downloaded from GenBank or J. H. Nitta specimen collection numbers
for sequences newly obtained by this study. Numbers at nodes indicate local
support values computed with the Shimodaira–Hasegawa test.
`r ape::Ntip(grammitid_tree)` tips; `r ape::Nnode(grammitid_tree)` internal nodes.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

nectandra_DNA_accessions.csv: DNA accession numbers and specimen accession numbers
of pteridophytes from the Nectandra Cloud Forest Reserve, Costa Rica.

Number of variables: `r ncol(nectandra_dna)`

Number of cases/rows: `r nrow(nectandra_dna)`

Variable list:

  - genomic_id: Genomic accession number assigned during DNA extraction, of the
	form "JNG" plus a four-digit number. Unique values.

  - specimen_id: Specimen accession number assigned to each specimen in
	nectandra_specimens.csv. Integer (not unique).

Missing data codes: no missing data.

Specialized formats or other abbreviations used: none.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

```{r rbcL-fasta-nums}
rbcL_fasta_min <- purrr::map_dbl(nectandra_rbcL_raw, length) %>% min
rbcL_fasta_max <- purrr::map_dbl(nectandra_rbcL_raw, length) %>% max
rbcL_fasta_mean <- purrr::map_dbl(nectandra_rbcL_raw, length) %>% mean %>% round
```

nectandra_rbcL.fasta: Newly generated rbcL sequences of pteridophytes from
the Nectandra Cloud Forest Reserve, Costa Rica in FASTA format. All species
included occur at the Nectandra Cloud Forest Reserve, Costa Rica; a small
number of sequences are from specimens collected elsewhere. Sequence names
correspond to 'genomic_id' in nectandra_DNA_accessions.csv. 
`r length(nectandra_rbcL_raw)` sequences; shortest sequence `r rbcL_fasta_min`
bp; longest sequence `r rbcL_fasta_max` bp; mean sequence length 
`r rbcL_fasta_mean` bp. Exported from Geneious project folder "Clean Sporos
Trimmed Genbank Submission" (raw Geneious project file not included in this
dataset).

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

```{r alignment-nums}
# Extract relevant lines from iqtree log
align_stats <-
  magrittr::extract(
    iqtree_log,
    iqtree_log %>% str_detect("Alignment has [:digit:]+ sequences"))
```

nectandra_rbcL.phy: Aligned rbcL sequences of pteridophytes from the
Nectandra Cloud Forest Reserve, Costa Rica in PHYLIP format.
`r dim(nectandra_rbcL)[[1]]` sequences; `r dim(nectandra_rbcL)[[2]]` bp;
`r get_number(align_stats, "informative sites")` parsimony-informative sites.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

nectandra_rbcL.treefile: Phylogenetic tree of pteridophytes from the
Nectandra Cloud Forest Reserve, Costa Rica in Newick format inferred with
IQTREE [@Nguyen2015]. Values at each node indicate SH-aLRT support (%) / UFboot support
(%). `r ape::Ntip(rbcL_tree)` tips; `r ape::Nnode(rbcL_tree)` internal nodes.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

nectandra_specimens.csv: Specimen data of pteridophytes from the Nectandra
Cloud Forest Reserve, Costa Rica collected by Joel Nitta. Formatting UTF-8.

Number of variables: `r ncol(nectandra_specimens)`

Number of cases/rows: `r nrow(nectandra_specimens)`

Variable list:

  - specimen_id: Unique specimen identification number (integer).

  - specimen: Voucher specimen number.

  - genus: Genus

  - specific_epithet: Specific epithet.

  - infraspecific_rank: Infraspecific rank.

  - infraspecific_name: Infraspecific name.

  - certainty: Degree of taxonomic certainty if not completely certain.

  - species: Species (genus plus specific epithet).

  - taxon: Species plus infraspecific name.

  - scientific_name: Taxon plus scientific author.

  - country: Country of origin.

  - locality: General area of collection.

  - site: Specific site where collected.

  - observations: Observations about specimen.

  - elevation: Elevation in m.

  - latitude: Latitude in decimal-degrees.

  - longitude: Longitude in decimal-degrees.

  - collector: Name of collector.

  - other_collectors: Names of other collectors if present.

  - herbaria: Codes of herbaria where voucher specimens are lodged.

  - date_collected: Date collected in YYYY-MM-DD format.

Missing data codes: Missing or non-applicable data have no values
(nothing entered between commas in the CSV file).

Specialized formats or other abbreviations used: Herbaria codes follow
Index Herbariorum [@Thiers2020], except for "Nectandra", which indicates
the private herbarium at the Nectandra Cloud Forest Reserve.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

ppgi_taxonomy.csv: Taxonomic system of Pteridophyte Phylogeny Group I (2016) for
pteridophytes at the genus level and above. Updated with one new genus (Hiya).

Number of variables: `r ncol(ppgi)`

Number of cases/rows: `r nrow(ppgi)`

Variable list:

  - class: Class.

  - order: Order.

  - suborder: Suborder.

  - family: Family.

  - subfamily: Subfamily.

  - genus: Genus.

Missing data codes: Non-applicable data have no values (nothing entered between
commas in the CSV file).

Specialized formats or other abbreviations used: none.

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-

REFERENCES
